---
title: "Story 2"
author: "Tony Fraser"
date: "19/Sept/2024"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{graphicx}
  - \usepackage{mdframed}
---

# Requesting API Credit
I did full integrations with FRED and BLS to power this chart. You can this code as well if you store your keys in the same place. 

# Assignment 

The Federal Reserve's mandate from Congress is to control inflation and to maintain low unemployment. These seem to be contradictory objectives. Has the FED been able to fulfill the mandate given to it by Congress?

\newpage

```{r build_dataset, message=FALSE, warning=FALSE, echo=FALSE}
library(blsAPI)
library(fredr) 
library(dplyr)
library(jsonlite)
library(tibble)
library(lubridate)

cpi_series_id <- 'CUSR0000SA0'
# CPI for All Urban Consumers (CPI-U), seasonally adjusted.
# measures inflation and the cost of living over time.

unemp_series_id <- 'LNS14000000'
# unemployment rate for peple over 16, seasonally adjusted.

start_year <- 1998
end_year <- 2024


get_bls_data <- function(series_id, start_year, end_year) { 
  api_key <- readLines("../keys/nogit-bls-api-key-32charslong")
  combined_data <- tibble()  # <- empty struct for partitioning 
  
  # api limit -> loop through the years in 20-year chunks 
  for (year in seq(start_year, end_year, by = 20)) {
    chunk_end_year <- min(year + 19, end_year)
    payload <- list(
      'seriesid' = series_id,
      'startyear' = as.character(year),
      'endyear' = as.character(chunk_end_year),
      'registrationkey' = api_key
    )
    
    # call the api
    response <- blsAPI(payload = payload, api_version = 2)
    response_json <- fromJSON(response, flatten = TRUE)
    
    # Extract and straighten
    chunk_data <- response_json$Results$series$data[[1]] %>%
      select(year, period, periodName, value) %>%
      mutate(value = as.numeric(value)) %>%  
      rename(month = periodName) %>%
      mutate(
        # Create a date using year-month-day format
        date = ymd(paste(year, month, "01", sep = "-"))  
      ) %>% 
      select('date', 'year', 'month', 'value') %>%
      as_tibble()
    
     # Combine with root partition
    combined_data <- bind_rows(combined_data, chunk_data)
  }    
  return(combined_data)
}

get_fred_data <- function(series_id, start_year, end_year) { 
  fredr_set_key(readLines("../keys/nogit-fred-api-key-32charslong"))
  combined_data <- tibble()  # <- empty struct for partitioning 

  # api limit -> loop through the years in 20-year chunks 
  for (year in seq(start_year, end_year, by = 20)) {
    chunk_end_year <- min(year + 19, end_year)
    chunk_data <- fredr(
      series_id = series_id,
      observation_start = as.Date(paste0(year, "-01-01")),
      observation_end = as.Date(paste0(chunk_end_year, "-12-31"))
    ) %>%
    select(date, value) %>%  # Select the columns you want
    mutate(year = format(as.Date(date), "%Y")) %>% 
    rename(federal_funds_rate = value) %>% 
   mutate(
    year = format(date, "%Y"),      # Extract year from the date
    month = format(date, "%B")      # Extract full month name from the date
    ) %>%
    select('date', 'year', 'month', 'federal_funds_rate') %>%
    as_tibble()
    
    # Combine with root partition
    combined_data <- bind_rows(combined_data, chunk_data)
  }
  return(combined_data)
}

## all these data sets are roughly the same structure. 
cpi_data <- get_bls_data(cpi_series_id , start_year, end_year) %>% 
  rename(consumer_price_index = value)

unemp_data <- get_bls_data(unemp_series_id ,start_year, end_year) %>%
  rename(unemployment_rate = value)

fed_data <- get_fred_data("FEDFUNDS", start_year, end_year)
```
```{r merge_dataframes, message=FALSE, warning=FALSE, echo=FALSE}
# If the reader of this code is new to api work, this is you do it, 
# build the date sequence then left join. 

start_date <- as.Date(paste0(start_year, "-01-01"))
end_date <- min(as.Date(paste0(end_year, "-12-31")), floor_date(Sys.Date(), "month") - months(1))

date_seq <- tibble(
  date = seq.Date(from = start_date, to = end_date, by = "month"),
  year = year(date),
  month = format(date, "%B")  # Get full month name for consistency with other data
)

merged_data <- date_seq %>%
  left_join(cpi_data, by = "date") %>%
  left_join(unemp_data, by = "date") %>%
  left_join(fed_data, by = "date") %>%
  select(
    date,
    year = year.x,  month = month.x,    # reduce the date columns
    consumer_price_index,
    unemployment_rate,
    federal_funds_rate
  )

# Data looks like this. 
#    date        year month     consumer_price_index unemployment_rate
#    <date>     <dbl> <chr>                    <dbl>             <dbl>
#  1 1998-01-01  1998 January                   162                4.6
#  2 1998-02-01  1998 February                  162                4.6
#  3 1998-03-01  1998 March                     162                4.7
#  4 1998-04-01  1998 April                     162.               4.3
#  5 1998-05-01  1998 May                       163.               4.4
#  6 1998-06-01  1998 June                      163.               4.5
#  7 1998-07-01  1998 July                      163.               4.5
#  8 1998-08-01  1998 August                    163.               4.5
#  9 1998-09-01  1998 September                 164.               4.6
# 10 1998-10-01  1998 October                   164.               4.5
# # ℹ 310 more rows
```
```{r build_plot2, fig.width=8, fig.height=12, echo=FALSE, message=FALSE, warning=FALSE}

## This is the base R chart. Man what a pain.

png("nogit_tmp_plot.png", width = 800, height = 1200, res = 100)

# Set up the layout with outer margins and adjust margins
par(mfrow = c(2, 1), oma = c(6, 4, 6, 4), mar = c(4, 4, 2, 2))

# # a sequence of years, by five, better to 
# years <- seq(from = as.Date("1998-01-01"), to = as.Date("2024-12-31"), by = "5 years")

years <- seq(from = as.Date(paste0(start_year, "-01-01")), 
             to = as.Date(paste0(end_year, "-12-31")), 
             by = "5 years")

# Define event dates for 9/11, Financial Crisis, COVID-19
event_dates <- as.Date(c("2001-09-11", "2008-09-15", "2020-03-11"))
event_colors <- c("red", "purple", "orange")  # Colors for the events

# Top plot: CPI with gray dotted lines in 5-year increments
plot(merged_data$date, merged_data$consumer_price_index, type = "l", col = "blue",
     xlab = "", ylab = "Consumer Price Index (Dollars)", 
     main = "Consumer Price Index (CPI) Over Time", xaxt = 'n')
abline(v = years, col = "lightgray", lty = "dotted")  # dotted 5-year lines
for (i in seq_along(event_dates)) {
  abline(v = event_dates[i], col = event_colors[i], lty = 2)  # Event lines in both charts
}

# Bottom plot: unemployment and federal funds 
plot(merged_data$date, merged_data$unemployment_rate, type = "l", col = "red",
     xlab = "Date", ylab = "Rate (%)",
     ylim = range(c(merged_data$unemployment_rate, merged_data$federal_funds_rate)),
     main = "Unemployment Rate and Federal Funds Rate Over Time", 
     xaxt = 'n')  # Suppress default x-axis
lines(merged_data$date, merged_data$federal_funds_rate, type = "l", col = "green")
abline(v = years, col = "lightgray", lty = "dotted")   # dotted 5-year lines
for (i in seq_along(event_dates)) {
  abline(v = event_dates[i], col = event_colors[i], lty = 2)  # Same event lines
}

# Manually add the x-axis ticks for the years variable
axis(1, at = years, labels = format(years, "%Y"), las = 1)

# Add the floating legend between the two plots, and spend hours making it fit in the right place.
par(xpd = NA)  # Allow drawing outside the plot area
legend("top", inset = c(0, -0.24), legend = c("9/11", "Financial Crisis", "COVID"),
       col = event_colors, lty = 2, cex = 0.8, horiz = TRUE)

# Add a legend to the bottom plot for Unemployment Rate and Federal Funds Rate
legend("topright", legend = c("Unemployment Rate", "Federal Funds Rate"),
       col = c("red", "green"), lty = 1, cex = 0.8)

dev.off()  # Write to file
```

\begin{mdframed}[linecolor=black, linewidth=3pt]
\begin{center}
\includegraphics[width=0.8\textwidth]{nogit_tmp_plot.png}
\end{center}
\end{mdframed}
