---
title: "Multiple linear regression"
format: 
  pdf:
    latex-engine: xelatex # or pdflatex or lualatex
    keep-tex: true
    preamble: |
      \usepackage{xcolor}
      \usepackage{framed}
---

\newcounter{exerciseNum}
\setcounter{exerciseNum}{0}

\newenvironment{exercise}[1][]
{
  \definecolor{myblue}{HTML}{1E2A38}
  \colorlet{shadecolor}{myblue}
  \begin{shaded}
  \stepcounter{exerciseNum} % Increment the exercise counter
  \ifx&#1& % Check if the optional argument is provided
    \noindent\textcolor{white}{\textbf{Exercise \theexerciseNum}}
  \else
    \noindent\textcolor{white}{\textbf{Exercise \theexerciseNum: #1}}
  \fi
  \par\noindent\hrule\vspace{0.5\baselineskip}
}
{
  \par\vspace{0.5\baselineskip}\noindent\hrule\end{shaded}
}



```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate the course and the instructor anonymously. However, the use of these student evaluations as an indicator of course quality and teaching effectiveness is often criticized because these measures may reflect the influence of non-teaching related characteristics, such as the physical appearance of the instructor. The article titled, "Beauty in the classroom: instructors' pulchritude and putative pedagogical productivity" by Hamermesh and Parker found that instructors who are viewed to be better looking receive higher instructional ratings. 

Here, you will analyze the data from this study in order to learn what goes into a positive professor evaluation.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
library(gridExtra)
library(corrplot)
```

This is the first time we're using the `GGally` package. You will be using the `ggpairs` function from this package later in the lab.

### The data

The data were gathered from end of semester student evaluations for a large sample of professors from the University of Texas at Austin. In addition, six students rated the professors' physical appearance. The result is a data frame where each row contains a different course and columns represent variables about the courses and professors. It's called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and some numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

\begin{exercise}[observational or an experiment?]
\textcolor{white}{Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.}
\end{exercise}

__This was an observational study. It would be an experiment if the same instructor came in all frumpy dressed and not put together one semester, and then the next semester he came totally put together, which is not what happened. Yes I think it is possible. It might be better to infer beauty leads to higher course evaluations, but as is beauty definitely leads to differences in course evaluations.__


\begin{exercise}[distribution of score]
\textcolor{white}{Describe the distribution of score. Is the distribution skewed? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?}
\end{exercise}

```{r}
ggplot(evals, aes(x = score)) + 
  geom_histogram(binwidth = 0.20, fill = "blue", color = "black") +
  ggtitle("Distribution of Scores") +
  xlab("Score") +
  ylab("Frequency") +
  xlim(0, 5) # Set the limits of the x-axis
```

_These results definitely skew toward higher ratings. It's interesting that so few ratings were less than the middle number of 2.5._

\begin{exercise}[Other variables]
\textcolor{white}{Excluding score, select two other variables and describe their relationship with each other using an appropriate visualization.}
\end{exercise}

```{r}
boxplot_gender <- ggplot(evals, aes(x = gender, y = score, fill = gender)) +
  geom_boxplot() +
  labs(title = "Scores by Gender", x = "", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) 

boxplot_ethnicity <- ggplot(evals, aes(x = ethnicity, y = score, fill = ethnicity)) +
  geom_boxplot() +
  labs(title = "Scores by Ethnicity", x = "", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) 

grid.arrange(boxplot_gender, boxplot_ethnicity, ncol = 2)
```

__Wow. So beautiful, white, and male all mean you're going to have higher ratings if you are an instructor. I guess teachers have the same problems we have, we all gotta hit the gym if we want the satisfaction scores.__

\newpage

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers are evaluated more favorably. Let's create a scatterplot to see if this appears to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of observations in the data frame with the approximate number of points on the scatterplot. Is anything awry?

\newpage

\begin{exercise}[geom\_point versus gemo\_gitter]
\textcolor{white}{Replot the scatterplot, but this time use geom\_jitter as your layer. What was misleading about the initial scatterplot?}
\end{exercise}


```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

__Geom Jitter shows where the points bunch up. Geom_point seems to be more about showing clean points and lines and doesn't communicate clustering as effectively.__

\newpage

\begin{exercise}[lm fit for beauty to score]
\textcolor{white}{Let's see if the apparent trend in the plot is something more than natural variation. Fit a linear model called m\_bty to predict average  professor score by average beauty rating. Write out the equation for the linear model and interpret the slope. Is average beauty score a statistically significant predictor? Does it appear to be a practically significant predictor?}
\end{exercise}

```{r}
m_bty <- lm(score ~ bty_avg, data = evals)
summary(m_bty)
```
$$
\text{score} = 0.06664 \times \text{bty\_avg} + 3.88034
$$

__The P value is 5.08e-05, which is much less than common significance level of 0.05. That means it is statistically significant, and is unlikely to have occurred by chance.__

Add the line of the bet fit model to your plot using the following:
    
```{r scatter-score-bty_avg-line-se}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
```

The blue line is the model. The shaded gray area around the line tells you about the variability you might expect in your predictions. To turn that off, use `se = FALSE`.

```{r scatter-score-bty_avg-line}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```
\newpage
\begin{exercise}[Further evaluation]
\textcolor{white}{Use residual plots to evaluate whether the conditions of least squares regression are reasonable. Provide plots and comments for each one (see the Simple Regression Lab for a reminder of how to make these)}
\end{exercise}
```{r fig.height=5}
m_bty_gen_0 <- lm(score ~ bty_avg, data = evals)
par(mfrow=c(2,2)) 
plot(m_bty_gen_0)
```

**The key assumptions of a linear regression model are met, which support the model's reliability. Residuals vs Fitted: Shows no clear patterns, suggesting linearity and homoscedasticity. Normal Q-Q Plot: Points align well with the line, indicating normally distributed residuals. Scale-Location Plot: Even spread of residuals across fitted values confirms homoscedasticity. Residuals vs Leverage: No signs of influential outliers, indicating a stable model.**

\newpage
## Multiple linear regression
The data set contains several variables on the beauty score of the professor: individual ratings from each of the six students who were asked to score the physical appearance of the professors and the average of these six scores. Let's take a look at the relationship between one of these scores and the average beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average score is calculated using the individual scores. You can actually look at the relationships between all beauty variables (columns 13 through 19) using the following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of these variables to the model would not add much value to the model. In this application and with these highly-correlated predictors, it is reasonable to use the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score after you've accounted for the professor's gender, you can add the gender term into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```
\newpage
\begin{exercise}[model conditions]
\textcolor{white}{P\-values and parameter estimates should only be trusted if the conditions for the regression are reasonable. Verify that the conditions for this model are reasonable using diagnostic plots.}
\end{exercise}


```{r fig.height=7, fig.width=10}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
par(mfrow=c(2,2)) 
plot(m_bty_gen)
```
**The key assumptions of a linear regression model are met, which support the model's reliability. Residuals vs Fitted: Shows no clear patterns, suggesting linearity and homoscedasticity. Normal Q-Q Plot: Points align well with the line, indicating normally distributed residuals. Scale-Location Plot: Even spread of residuals across fitted values confirms homoscedasticity. Residuals vs Leverage: No signs of influential outliers, indicating a stable model. (Note: this is same statement as question #6) **

\newpage
\begin{exercise}[Still a valid predictor?]
\textcolor{white}{Is bty\_avg still a significant predictor of score? Has the addition of gender to the model changed the parameter estimate for bty\_avg?}
\end{exercise}

 **The coefficient of bty_avg increased to 0.07416. This suggests a stronger relationship between bty_avg and score when gender is also included in the model. The coefficient for gendermale indicates that male professors have an average score that is 0.17239 points higher than female professors, controlling for beauty average.**



Note that the estimate for `gender` is now called `gendermale`. You'll see this name change whenever you introduce a categorical variable. The reason is that R recodes `gender` from having the values of `male` and `female` to being an indicator variable called `gendermale` that takes a value of $0$ for female professors and a value of $1$ for male professors. (Such variables are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied by zero, leaving the intercept and slope form familiar from simple regression.

$$
  \begin{aligned}
    x &= y + z
  \end{aligned}
$$


$$
  \begin{aligned}
    \widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
                    &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg
  \end{aligned}
$$ 


```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

\newpage
\begin{exercise}[What is the equation of the line?]
\textcolor{white}{What is the equation of the line corresponding to those with color pictures? (*Hint:* For those with color pictures, the parameter estimate is multiplied by 1.) For two professors who received the same beauty rating, which color picture tends to have the higher course evaluation score?}
\end{exercise}

$$
\text{score} = (\text{4.06318}) + (\text{0.05548}) \times \text{Beauty Average} + (\text{-0.16059}) \times 1
$$


**For the second part of the question, I'm not exactly sure what it is asking, and you can sort of see in the graph that black and white photos seem to do well, but when you look at raw data it doesn't look as convincing.**

```{r}
avg_scores <- evals %>%
  group_by(bty_avg, pic_color) %>%
  summarise(avg_score = mean(score, na.rm = TRUE))
print(avg_scores, n=10) 
```


The decision to call the indicator variable `gendermale` instead of `genderfemale` has no deeper meaning. R simply codes the category that comes first alphabetically as a $0$. (You can change the reference level of a categorical variable, which is the level that is coded as a 0, using the`relevel()` function. Use `?relevel` to learn more.)


\begin{exercise}[use rank]
\textcolor{white}{Create a new model called m\_bty\_rank with gender removed and rank added in. How does R appear to handle categorical variables that have more than two levels? Note that the rank variable has three levels, teaching, tenure track, tenured.}
\end{exercise}


```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
m_bty_rank
```
**It incorporates another of the values within the variable. What I don't quite understand is that it has statistics for two of the three values. Like in gender, there is also gender:female, not just gender:male. The model doesn't show gender:female**

```{txt}
> table(evals$gender) 

female   male 
   195    268 
> lm(score ~ bty_avg + gender, data = evals)

Call:
lm(formula = score ~ bty_avg + gender, data = evals)

Coefficients:
(Intercept)      bty_avg   gendermale  
    3.74734      0.07416      0.17239  
```

The interpretation of the coefficients in multiple regression is slightly different from that of simple regression. The estimate for `bty_avg` reflects how much higher a group of professors is expected to score if they have a beauty rating that is one point higher *while holding all other variables constant*. In this case, that translates into considering only professors of the same rank with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, gender, ethnicity, language of the university where they got their degree, age, proportion of students that filled out evaluations, class size, course level, number of professors, number of credits, average beauty rating, outfit, and picture color.

\begin{exercise}[which variable doesn't matter?]
\textcolor{white}{Which variable would you expect to have the highest p-value in this model? Why? *Hint:* Think about which variable would you expect to not have any   association with the professor score.}
\end{exercise}

**I looked it up, it's number of professors.**

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

\begin{exercise}[check your answer?]
\textcolor{white}{Check your suspicions from the previous exercise. Include the model output in your response.}
\end{exercise}


```{text}
> m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full) 

Call:
lm(formula = score ~ rank + gender + ethnicity + language + age + 
    cls_perc_eval + cls_students + cls_level + cls_profs + cls_credits + 
    bty_avg + pic_outfit + pic_color, data = evals)

Residuals:
     Min       1Q   Median       3Q      Max 
-1.77397 -0.32432  0.09067  0.35183  0.95036 

Coefficients:
                        Estimate Std. Error t value Pr(>|t|)    
(Intercept)            4.0952141  0.2905277  14.096  < 2e-16 ***
ranktenure track      -0.1475932  0.0820671  -1.798  0.07278 .  
ranktenured           -0.0973378  0.0663296  -1.467  0.14295    
gendermale             0.2109481  0.0518230   4.071 5.54e-05 ***
ethnicitynot minority  0.1234929  0.0786273   1.571  0.11698    
languagenon-english   -0.2298112  0.1113754  -2.063  0.03965 *  
age                   -0.0090072  0.0031359  -2.872  0.00427 ** 
cls_perc_eval          0.0053272  0.0015393   3.461  0.00059 ***
cls_students           0.0004546  0.0003774   1.205  0.22896    
cls_levelupper         0.0605140  0.0575617   1.051  0.29369    
cls_profssingle       -0.0146619  0.0519885  -0.282  0.77806    
cls_creditsone credit  0.5020432  0.1159388   4.330 1.84e-05 ***
bty_avg                0.0400333  0.0175064   2.287  0.02267 *  
pic_outfitnot formal  -0.1126817  0.0738800  -1.525  0.12792    
pic_colorcolor        -0.2172630  0.0715021  -3.039  0.00252 ** 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.498 on 448 degrees of freedom
Multiple R-squared:  0.1871,    Adjusted R-squared:  0.1617 
F-statistic: 7.366 on 14 and 448 DF,  p-value: 6.552e-14

```

\begin{exercise}[ethnicity]
\textcolor{white}{Interpret the coefficient associated with the ethnicity variable.}
\end{exercise}

**Ethnicty isn't a significant varable.**

\begin{exercise}[ethnicity]
\textcolor{white}{Drop the variable with the highest p\-value and re fit the model. Did the coefficients and significance of the other explanatory variables change? (One of the things that makes multiple regression interesting is that coefficient estimates depend on the other variables that are included inthe model.) If not, what does this say about whether or not the dropped variable was collinear with the other explanatory variables?}
\end{exercise}

```{text}
> m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level  + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full) 

> m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level  + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full) 

Call:
lm(formula = score ~ rank + gender + ethnicity + language + age + 
    cls_perc_eval + cls_students + cls_level + cls_credits + 
    bty_avg + pic_outfit + pic_color, data = evals)

Residuals:
    Min      1Q  Median      3Q     Max 
-1.7836 -0.3257  0.0859  0.3513  0.9551 

Coefficients:
                        Estimate Std. Error t value Pr(>|t|)    
(Intercept)            4.0872523  0.2888562  14.150  < 2e-16 ***
ranktenure track      -0.1476746  0.0819824  -1.801 0.072327 .  
ranktenured           -0.0973829  0.0662614  -1.470 0.142349    
gendermale             0.2101231  0.0516873   4.065 5.66e-05 ***
ethnicitynot minority  0.1274458  0.0772887   1.649 0.099856 .  
languagenon-english   -0.2282894  0.1111305  -2.054 0.040530 *  
age                   -0.0089992  0.0031326  -2.873 0.004262 ** 
cls_perc_eval          0.0052888  0.0015317   3.453 0.000607 ***
cls_students           0.0004687  0.0003737   1.254 0.210384    
cls_levelupper         0.0606374  0.0575010   1.055 0.292200    
cls_creditsone credit  0.5061196  0.1149163   4.404 1.33e-05 ***
bty_avg                0.0398629  0.0174780   2.281 0.023032 *  
pic_outfitnot formal  -0.1083227  0.0721711  -1.501 0.134080    
pic_colorcolor        -0.2190527  0.0711469  -3.079 0.002205 ** 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.4974 on 449 degrees of freedom
Multiple R-squared:  0.187,     Adjusted R-squared:  0.1634 
F-statistic: 7.943 on 13 and 449 DF,  p-value: 2.336e-14

```

**Removing that variable doesn't seem to make much of a differnece.**



\begin{exercise}[ethnicity]
\textcolor{white}{Using backward\-selection and p\-value as the selection criterion, determine the best model. You do not need to show all steps in your answer, just the output for the final model. Also, write out the linear model for predicting score based on the final model you settle on.}
\end{exercise}

```{r}
m_backwards <- lm(score ~ gender + ethnicity + language + age + cls_perc_eval + cls_credits + bty_avg + pic_outfit + pic_color, data = evals) 
summary(m_backwards)

```` 

\begin{exercise}[verify the conditions]
\textcolor{white}{Verify that the conditions for this model are reasonable using diagnostic plots.}
\end{exercise}


```{r fig.height=6}
par(mfrow=c(2,2)) 
plot(m_backwards)

```

**Looks good, but also very interesting. The graphs are a little gither than the previous versions.**


\begin{exercise}[texas]
\textcolor{white}{The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression? }
\end{exercise}

**To use linear regression, you have to have independence between records. Though each course is an independent record, if we're looking at individual professor beauty scores that might not be as independent as necessary.**



\begin{exercise}[who gets good grades?]
\textcolor{white}{Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.}
\end{exercise}

**Usually male, beautiful and younger.**


\begin{exercise}[who gets good grades?]
\textcolor{white}{Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?}
\end{exercise}

**Of course it's the same. It isn't just teachers, it's car dealers, food servers, pretty much anyobody. Beautiful people get higher rankings. It's some kind of Darwin trait we all have burned into us.**

