---
title: "Assignment 3: Strings and Regexes"
author: "By Tony Fraser, CUNY MSDS Student"
date: "17/Sept/2023"
output:
  pdf_document: default
  latex_engine: xelatex # or pdflatex or lualatex
monofontoptions: 
  - Scale=0.55
editor_options: 
  chunk_output_type: console
---
##  Assignment:
Use Quarto and markdown to asnwer the following questions.

## Question 1:
_Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset used for this [study](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major), and provide code that identifies the majors that contain either "DATA" or "STATISTICS"_

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(gt)
fivethirtyeight <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/"
m_list <- "college-majors/majors-list.csv"
filter_for <- ".*data.*|*.statistics*."
majors <- read_csv(paste(fivethirtyeight, m_list, sep="")) |> 
    filter(grepl(filter_for, Major, ignore.case = TRUE) | 
           grepl(filter_for, Major_Category, ignore.case = TRUE)) |>
   rename(ID = FOD1P)
majors |> 
    gt() %>%
    tab_options(table.font.size = px(10)) %>% 
    tab_style(style = cell_text(size = 'smaller'), locations = cells_body())
```
\newpage
## Question 2
Write code that transforms input data like 
_[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe" ... _ and writes output data like _
c("bell pepper", "bilberry", "blackberry", "blood orange", 
"blueberry", "cantaloupe", "chili ..._ Or basically, read a string input, puts it into a vector, then prints it out so it looks like a vector.


```{r message=FALSE, warning=FALSE}
library(magrittr)
library(stringr)

to_vector <- function(input_string) {
  cleaned_vector <- input_string %>%
    gsub("\\[\\d+\\]|\\\"", "", .) %>% # remove []"
    gsub(" {2,}|\n", ",", .) %>%       # replace 2+ \s with ,
    str_split(",") %>%                 # split str to list w/ vetor
    `[[`(1) %>%                        # same as .[[1]]%>%, extract first element
    str_trim()                         # trim whitespace off elements
  return(cleaned_vector)
}

to_string <- function(v) {
  var_name <- deparse(substitute(v))
  v %>%
   lapply(function(x) paste0("\"", x, "\"")) %>%   # surround elements with quotes
   unlist() %>%                                    # convert list to a character vector
   paste(collapse = ", ") %>%                      # collapse to string
   paste(var_name, " <- c(", ., ")", sep = "") %>% # finish return string
  return(v)
}

input_string <- paste(
    '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"',
    '[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"',
    '[9] "elderberry"   "lime"         "lychee"       "mulberry"',
    '[13] "olive"        "salal berry"',
    sep = "\n"
)

jack <- to_vector(input_string)
cat(to_string(jack)) # use cat instead of print, doesn't print out "'s.\newpage
```

\newpage
## Question 3
Describe what these regexes match

#### 1. (.)\1\1
This matches one character three times in a row, like ppQQ:
```r
> regex <- "(.)\\1\\1"
> matches <- str_match_all("tony aaabb", regex)
> matches[[1]][ ,1]
[1] "aaa"
```
#### 2. (.)(.)\\2\\1
Matches two characters and their reverse, like abba!
```r
regex <-"(.)(.)\\2\\1"
matches <- str_match_all("dancing queen abba") 
> matches[[1]][ ,1]
[1] "abba"
```
#### 3. (..)\1
Matches two characters right after each other, like lulu.  
```r
regex <-"(..)\\1"
matches <- str_match_all("eat me some hohos", regex)
> matches[[1]][ ,1] 
[1] "hoho"
```
#### 4. (.).\\1.\\1
Think R?R?R, or a triple decker sandwich of characters with anything in between.  The first (.) picks what those 3 matching characters will be, and the periods in there mean "anything."

```r
> matches <- str_match_all("rxrxrx rarbrc r?r*r", regex) 
> matches
[[1]]
     [,1]    [,2]
[1,] "rxrxr" "r" 
[2,] "rarbr" "r" 
[3,] "r?r*r" "r" 

```
#### 5. (.)(.)(.).*\\3\\2\\1
Think of this one like a sandwich also, three characters as the bread, and whatever you want in between. Except the ending bread slice needs to be backwards.
```r
> regex<- "(.)(.)(.).*\\3\\2\\1"
> matches <- str_match_all("abc what's this cba  def tony fed", regex) 
> matches
[[1]]
     [,1]                  [,2] [,3] [,4]
[1,] "abc what's this cba" "a"  "b"  "c" 
[2,] "def tony fed"        "d"  "e"  "f" 
```

## Question 4, build regexes that match:
#### 4.1 Words that start and end with the same character
\\b means "word boundary." So in the word "word", the boundaries are w and b. "./*"  means anything, and \\1\\b means it looks at the other word boundary and sees if it matches the first group. 

```r
> regex <- "(\\b.).*\\1\\b"  
> matches <- str_extract_all("bob tot tonyfraserwashere", regex) 
> matches
[[1]]
[1] "bob"   " tot "
```
Note, snippit 4.1 searches for all words within a string, not just a string exact match like `"^(.).*\1$"`

#### 4.2 Words contain a repeated pairs of letters
- like church contains ch repeated twice
* `"\\b(\\w*(\\w{2,})\\w*\\2\\w*)\\b"` is the regex we are going to use. 
* \\w means "word", which is shorthand for this. [a-zA-Z0-9_]. 
* The outer \b's are for word boundaries, meaning we are looking for groups of characters surrounded by spaces, characters, newlines, etc.  
* The first capture group is here, `(\\w*(\\w{2,})\\w*\\2\\w*)` it's the entire word. 
* The second capture group is `\\w{2,})`, and this is what \\2 looks to see if there is a repeat of. If \\2 exists, the first capture group matches.  

```r
> regex <- "\\b(\\w*(\\w{2,})\\w*\\2\\w*)\\b"
matches <- str_extract_all("church goer and chacha dancers", regex)  
> matches
[[1]]
[1] "church" "chacha"
```

#### 4.3 Words that contain one letter repeated three times
- like eleven with 3 e's.
* `"\\b\\w*(\\w)\\w*\\1\\w*\\1\\w*\\b"` is the regex we are going to use, and is very similar to the last example. 
* \\b matches words
* the first \\* mean pretty much any character
* the (\\w) is the important capture group, and the next two \1's are repeats of that capture 
* the \\w*'s between the \\1's mean any word character (or no character) 
```r
> regex <- "\\b\\w*(\\w)\\w*\\1\\w*\\1\\w*\\b"
> matches <- str_extract_all("eleven twelve altavista speaker", regex) 
> matches
[[1]]
[1] "eleven"    "altavista"

```